---
title: "220-week-03"
author: "Bob Rudis"
date: "1/29/2017"
output:
  html_document:
    toc: true
    code_download: true
    theme: simplex
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE, 
  warning = FALSE
)
```

```{r libraries_we_need}
library(knitr)
library(readxl)
library(tidyverse)
library(ggthemes)
library(mapproj)
```

## Exploring Pipeline Incidents

Using data from <http://www.phmsa.dot.gov/pipeline/library/data-stats/flagged-data-files>

```{r get_data}
URL <- "http://www.phmsa.dot.gov/staticfiles/PHMSA/DownloadableFiles/Pipeline/PHMSA_Pipeline_Safety_Flagged_Incidents.zip"
fil <- "data/pipelines.zip"
if (!file.exists(fil)) {
  download.file(URL, fil)
  unzip(fil, exdir="data")
}
data_file <- "data/hl2010toPresent.xlsx"
```

Read in the data:

```{r load_data}
pipelines <- read_excel(data_file, "hl2010toPresent", col_names=TRUE)
```

```{r eda, eval=FALSE, echo=FALSE}
# take a look at the strucure
glimpse(pipelines)
```

### Significant incidents

We want to identify if record meets the significant criteria or not:

- fatality
- injury
- fire
- explosion
- total property damage $50K or more in 1984 dollars
- non-HVL loss >= 50bbls
- HVL loss >= 5bbls

We'll do this by year.

```{r signigicant_incidents}
count(pipelines, IYEAR, SIGNIFICANT) %>% 
  arrange(IYEAR) %>% 
  spread(SIGNIFICANT, n) %>% 
  rename(`Incident year`=IYEAR,
         `Not Significant`=NO, 
         `Significant`=YES) -> sig_ind_by_year
```

`r kable(sig_ind_by_year)`

```{r sig_ind_plot}
gather(sig_ind_by_year, is_sig, ct, -`Incident year`) %>% 
  ggplot(aes(`Incident year`, ct, group=is_sig)) +
  geom_line(aes(color=is_sig)) +
  labs(title="Significant pipeline incidents by year (2010-2016)") +
  theme_bw() +
  theme(legend.position="bottom")
```

```{r sig_ind_plot_diff, fig.height=2}
mutate(sig_ind_by_year, diff=`Not Significant`- `Significant`) %>%
  ggplot(aes(`Incident year`, diff)) +
  geom_segment(aes(xend=`Incident year`, yend=0)) +
  labs(title="[Difference] Significant pipeline incidents by year (2010-2016)") +
  theme_bw() +
  theme(panel.grid.major.x=element_blank()) +
  theme(panel.grid.minor.x=element_blank())
```

### How about Serious (had a fatality or injury)?

```{r serious}
count(pipelines, IYEAR, SERIOUS) %>% 
  spread(SERIOUS, n) %>% 
  kable()
```

### How about Shutdowns?

```{r shutdown}
count(pipelines, IYEAR, SHUTDOWN_DUE_ACCIDENT_IND, sort=TRUE) %>%   filter(!is.na(SHUTDOWN_DUE_ACCIDENT_IND)) %>% 
  spread(SHUTDOWN_DUE_ACCIDENT_IND, n) %>% 
  kable()
```

### How about Spills?

```{r spills}
count(pipelines, IYEAR, SPILL_TYPE_CATEGORY, sort=TRUE) %>% 
  spread(SPILL_TYPE_CATEGORY, n) %>% 
  kable()
```

## Where does this stuff happen?

### By how serious the incident was (2013)

```{r where_2013}
filter(pipelines, IYEAR=="2013") %>% 
  ggplot(aes(LOCATION_LONGITUDE, LOCATION_LATITUDE)) +
  geom_point(aes(color=SERIOUS), size=0.75, alpha=1/2, shape=21) +
  coord_map("polyconic", xlim=c(-124.8, -66.9), ylim=c(24.3, 49.4)) +
  ggthemes::theme_map()
```

### By how Significant the incident was (2014)

```{r where_2014}
filter(pipelines, IYEAR=="2014") %>% 
  ggplot(aes(LOCATION_LONGITUDE, LOCATION_LATITUDE)) +
  geom_point(aes(color=SIGNIFICANT), size=0.75, alpha=1/2, shape=21) +
  coord_map("polyconic", xlim=c(-124.8, -66.9), ylim=c(24.3, 49.4)) +
  ggthemes::theme_map()
```

### Cumulative — 2010:2016 (Significant)

```{r where_cumulative}
pipelines %>% 
  ggplot(aes(LOCATION_LONGITUDE, LOCATION_LATITUDE)) +
  geom_point(aes(color=SIGNIFICANT), size=0.75, alpha=1/2, shape=21) +
  coord_map("polyconic", xlim=c(-124.8, -66.9), ylim=c(24.3, 49.4)) +
  ggthemes::theme_map()
```


```{r operator_most_injure}
filter(pipelines, INJURY_IND == "YES") %>%
  select(NAME, INJURE, IYEAR) %>%
  group_by(NAME, IYEAR) %>%
  summarise(totals = sum(INJURE)) -> operator_most_injures

years <- as.double(c("2010", "2011", "2012", "2013", "2014", "2015", "2016"))



operator_most_injures %>%
  filter(IYEAR == "2016") 

```

```{r prototype}
filter(pipelines, FATALITY_IND == "YES") %>%
  select(matches("FATAL|IYEAR"), -FATALITY_IND) %>%
  gather(IYEAR) -> a
colnames(a) <- c("year", "cat", "val")
a %>%
  arrange(year) %>%
  kable()
```

## Fatalities | Injuries by Type

```{r progress}
filter(pipelines, FATALITY_IND == "YES" | INJURY_IND == "YES") %>%
  select(matches("INJUR|IYEAR|FATAL"), -FATALITY_IND, -INJURY_IND) %>%
  gather(category, value, -IYEAR, na.rm = TRUE) %>%
  rename(year = IYEAR) %>%
  count(year, category, wt = value) -> pipelinetrans

separate(pipelinetrans, category, into = c("num", "type", "fat_or_inj"), sep = "_", fill = "left") %>%
  select(-num) %>%
  filter(!is.na(type)) -> fatlinjr



fatlinjr %>%
  group_by(year, type, fat_or_inj) %>%
  summarise(tot = sum(n)) %>%
  mutate(pct = (tot / sum(tot))*100) %>%
  arrange(fat_or_inj) -> summ_fatlinjr
kable(summ_fatlinjr)
```
```{r fatinjplots}
ggplot(summ_fatlinjr, aes(year, tot, fill = fat_or_inj)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

### Operator w/ Most Injuries

```{r q2}
filter(pipelines, FATALITY_IND == "YES" | INJURY_IND == "YES") %>%
  select(matches("INJUR|IYEAR|FATAL|OPERATOR_ID"), -FATALITY_IND, -INJURY_IND) %>%
  gather(category, value, -IYEAR, -OPERATOR_ID, na.rm = TRUE) %>%
  rename(year = IYEAR) %>%
  count(year, category, OPERATOR_ID, wt = value) %>%
  arrange(year)-> pipelinetrans2

separate(pipelinetrans2, category, into = c("num", "type", "fat_or_inj"), sep = "_", fill = "left") %>%
  select(-num) %>%
  filter(!is.na(type)) %>%
  count(year, OPERATOR_ID, fat_or_inj, wt = n) -> fatlinjr2
kable(fatlinjr2)

count(fatlinjr2, year, OPERATOR_ID, fat_or_inj, wt = n) %>%
  group_by(year, fat_or_inj) %>%
  filter(n == max(n)) %>%
  arrange(year) -> summ_fatlinjr2

filter(summ_fatlinjr2, fat_or_inj == "INJURIES") %>%
  kable()
```

### Operator w/ most Fatalities
```{r q3}

filter(summ_fatlinjr, fat_or_inj == "FATALITIES") %>%
  kable()
```

### Onshore or offshore?
```{r q4}
count(pipelines, ON_OFF_SHORE) %>%
  mutate(pct = (n / sum(n))*100) %>%
  kable()
```

### Incidents by Type of Pipe/Weld by Year and Failure Type

```{r q5}
pipelines %>%
  group_by(IYEAR) %>%
  count(PIPE_TYPE, WELD_SUBTYPE, FAILURE_TYPE, IYEAR) %>%
  ungroup() %>%
  arrange(IYEAR) %>%
  kable()

```
